Assessment of forestry road network projection methods using the “roads” package in R

### 2022-08-26 analysis/paper/paper.Rmd
new projection methods figure and add appendix with coarse results

```{r fig2, fig.height=5, fig.width=fig_widths["two"], fig.cap="Example of the different projection methods and sampling strategies."}
knitr::include_graphics(here("analysis/figures/projection methods figure.jpg"))
```

@@

Projection accuracy varied widely across the 5 metrics and two scales compared (Figure \@ref(fig:fig3)). Across all projection methods road projections were less accurate within cutblocks than in the whole TSA. Metrics that used a buffer around the roads, forestry and road disturbance footprint, showed closer agreement between the observed and projected roads than the road density and road presence metrics at both scales. For all methods the distance to road metric was underestimated within cutblocks but was similar to the observed across the whole TSA. The road presence and road density metrics were the most sensitive to pixel size and were less accurate when the coarse cost raster was used (reference appendix table for coarse resolution projections).

@@

As expected all the projection methods preformed worst with the road presence metric (F-Measure: `r perf_tbl_rng["Road presence"]`), while the forestry disturbance footprint performed best (F-Measure: `r perf_tbl_rng["Forestry disturbance footprint"]`) and road disturbance footprint only slightly worse (F-Measure: `r perf_tbl_rng["Road disturbance footprint"]`). For the road presence metric higher sampling densities led to somewhat higher F-measure scores caused by fewer false negatives and only slightly more false positives. For the forestry and road disturbance footprint metrics, the lower density projection methods had slightly higher F-measure scores due to fewer false positives and only slightly more false negatives.

@@

>>>>

```{r fig2, fig.height=5, fig.width=fig_widths["two"], fig.cap="Example of the different projection methods and sampling strategies."}
knitr::include_graphics(here("analysis/figures/projection_methods_figure.jpg"))
```

@@

Projection accuracy varied widely across the 5 metrics and two scales compared (Figure \@ref(fig:fig3)). Across all projection methods road projections were less accurate within cutblocks than in the whole TSA. Metrics that used a buffer around the roads, forestry and road disturbance footprint, showed closer agreement between the observed and projected roads than the road density and road presence metrics at both scales. For all methods the distance to road metric was underestimated within cutblocks but was similar to the observed across the whole TSA. The road density and road distance to road metrics were the most sensitive to pixel size and were less accurate when the coarse cost raster was used (\@ref(fig:figA1)).

@@

As expected all the projection methods preformed worst with the road presence metric (F-Measure: `r perf_tbl_rng["Road presence"]`), while the forestry disturbance footprint performed best (F-Measure: `r perf_tbl_rng["Forestry disturbance footprint"]`) and road disturbance footprint only slightly worse (F-Measure: `r perf_tbl_rng["Road disturbance footprint"]`). For the road presence metric higher sampling densities led to somewhat higher F-measure scores caused by fewer false negatives and only slightly more false positives. For the forestry and road disturbance footprint metrics, the F-measure was similar across projection methods.

@@

## Coarse resolution results

```{r get-data2}
data_path_drvd <- "analysis/data/derived_data/TSA27_1000"

# load csv
meanTable2 <- read.csv(here(data_path_drvd, "mean_table.csv")) %>% 
  mutate(resolution = "1000")

matchData2 <- read.csv(here(data_path_drvd, "agree_table.csv")) %>% 
  mutate(resolution = "1000")

meanTable <- meanTable %>% mutate(resolution = "100") %>% bind_rows(meanTable2)

matchData <- matchData %>% mutate(resolution = "100") %>% bind_rows(matchData2)

```{r figA1, fig.height=4, fig.width=fig_widths["mid"], fig.cap="Proportional difference of projected mean from observed mean road metrics using different projection methods across the entire Revelstoke TSA (“overall”) and within cutblocks (“cutblocks”) for the fine resolutiom (100 m) and coarse resolution (1000 m) cost rasters. Values greater than 0 indicate the projection overestimated the metric and negative values indicate an underestimation."}
# organize data for ggplot
meanTable_long <- meanTable %>%
  pivot_longer(-c(sampleType, sampleDens, areaMean, resolution),
               names_to = "metric", values_to = "response")%>%
  mutate(sampleType = ifelse(sampleType == sampleDens, NA_character_, 
                             sampleType),
         sampleDens = paste(sampleType, sampleDens) %>% 
           factor(levels = c("NA centroid",
                             "NA observed",
                             "random low sample density",
                             "regular low sample density",
                             "random high sample density",
                             "regular high sample density",
                             "NA klementQGIS"),
                  labels = c("Centroid", 
                             "Observed",
                             "Random\nlow density",
                             "Regular\nlow density",
                             "Random\nhigh density",
                             "Regular\nhigh density",
                             "Hardy QGIS")),
         metric = factor(metric,
                         levels = c("roadDensityMean", "roadPresenceMean",
                                    "distanceToRoadMean",
                                    "forestryDisturbanceMean",
                                    "roadDisturbanceMean"),
                         labels = c("Road\ndensity", "Road\npresence",
                                    "Distance\nto road",
                                    "Forestry\ndisturbance\nfootprint",
                                    "Road\ndisturbance\nfootprint")))

# get proportional difference from observed
prop_dif <- projected_values %>% 
  left_join(observed_values %>% select(areaMean, metric, resolution, response),
            by = c("areaMean", "metric", "resolution"), 
            suffix = c("_proj", "_obs")) %>% 
  mutate(prop_dif = (response_proj-response_obs)/response_obs,
         areaMean = factor(areaMean, labels = c("Cutblocks", "Overall")))

prop_dif %>% 
  ggplot(aes(x = sampleDens, prop_dif, fill = resolution))+
  geom_hline(aes(yintercept = 0), color = "grey")+
  geom_col(position = position_dodge2(preserve = "single",
                                      width = 0.75))+
  facet_grid(areaMean~ metric)+
  theme_classic()+
  scale_fill_manual(values = met.brewer(pal_nm))+
  theme(text = element_text(size = 10), axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        legend.position = "right",
        axis.text.x=element_text(size=11, angle=50, vjust=1, hjust=1))+
  coord_flip()

```{r figA2, fig.height=6, fig.width=fig_widths["mid"], fig.cap="Comparison of performance metrics for each projection method. F measure is the harmonic mean of precision and sensitivity, precision is the proportion of  predicted roads that are observed roads and sensitivity is the proportion of observed roads that were correctly predicted."}

matchData_sum <- matchData %>% ungroup() %>%
  mutate(agreement = factor(agreement,
                            levels = c("False negative", "False positive",
                                       "Agree roaded","Agree roadless",
                                       "Pre-existing roads")),
         sampleDens = paste(sampleType, sampleDens) %>%
           str_replace("1e-06", "low density") %>%
           str_replace("1e-05", "high density") %>%
           str_replace("centroid low density", "centroid") %>%
           str_replace("klementQGIS NA", "klementQGIS") %>%
           factor(levels = c("centroid",
                             "random low density",
                             "regular low density",
                             "random high density",
                             "regular high density",
                             "klementQGIS"),
                  labels = c("Centroid", 
                             "Random\nlow density",
                             "Regular\nlow density",
                             "Random\nhigh density",
                             "Regular\nhigh density",
                             "Hardy QGIS")),
         metric = factor(metric,
                         levels = c("roadPresence",
                                    "forestryDisturbance",
                                    "roadDisturbance"),
                         labels = c("Road presence",
                                    "Forestry disturbance\nfootprint",
                                    "Road disturbance footprint"))) %>%
  group_by(metric, sampleDens, agreement, resolution) %>%
  summarise(count = sum(count), .groups = "drop_last") %>%
  mutate(perc = count/sum(count)*100)

# performance table
perf_tbl <- matchData_sum %>% 
  pivot_wider(c(metric, sampleDens, resolution), names_from = agreement, 
              values_from = count) %>% 
  mutate(sensitivity = `Agree roaded`/(`Agree roaded` + `False negative`),
         precision = `Agree roaded`/(`Agree roaded` + `False positive`),
         F_measure = (2*precision*sensitivity)/(precision+sensitivity)) %>% 
  select(resolution, metric, sampleDens, sensitivity, precision, F_measure)

perf_tbl_rng <- perf_tbl %>% group_by(metric, resolution) %>% 
  summarise(rng_F = range(F_measure) %>% round(3) %>% paste0(collapse = " - "), .groups = "drop") %>% pull(rng_F, name = metric)

perf_tbl %>% 
  pivot_longer(c(sensitivity, precision, F_measure), names_to = "perf_meas",
               values_to = "value") %>% 
  mutate(perf_meas = factor(perf_meas,
                            labels = c("F measure", "Precision", "Sensitivity"))) %>% 
  ggplot(aes(sampleDens, value, fill = resolution))+
  geom_col(position = position_dodge2(preserve = "single",
                                      width = 0.75))+
  scale_fill_manual(values = met.brewer(pal_nm))+
  facet_grid(metric ~ perf_meas)+
  coord_flip()+
  ylim(0, 1)+
  theme_classic()+
  labs(y = "Performance", x = NULL)
